Charity During Crackdown: Analyzing the Impact of State Repression of NGOs on Philanthropy

### 2019-03-13 manuscript/manuscript.Rmd
Suparna edits

---
title: "Charity During Crackdown: Analyzing the Impact of State Repression of NGOs on Philanthropy"
short-title: Charity During Crackdown
author:
- name: Suparna Chaudhry
  affiliation: Christopher Newport University
  email: suparna.chaudhry@cnu.edu
  url: http://www.suparnachaudhry.com/
  twitter: SuparnaChaudhry
- name: Andrew Heiss
  affiliation: Brigham Young University
  email: andrew_heiss@byu.edu
  url: https://www.andrewheiss.com/
  twitter: andrewheiss
date: "`r format(Sys.time(), '%B %e, %Y')`"
published: Please do not cite without permission.
git-repo: https://github.com/andrewheiss/ngo-crackdowns-philanthropy
reference-section-title: References
toc: false
title-page: true
abstract: >-
  State crackdown on non-governmental organizations (NGOs), particularly through legal restrictions on funding, has become increasingly common. How do legal restrictions on NGOs impact patterns of private philanthropic giving and individual donor decision-making? With reduced funding for NGOs working on contentious issues, and an absence of philanthropic culture in many developing countries, many NGOs rely on funds from private donors in Western countries. Using a survey experiment, we explore how this crackdown changes donors' preferences based on the issue area and funding sources of the NGO. We find that private donors are responsive to the legal difficulties international NGOs face abroad and are more likely to donate to legally besieged privately funded human rights NGOs. Additionally, already-likely donors give substantially more to legally restricted NGOs working on humanitarian issues. We conclude by discussing the implications for the sustainability of NGOs working abroad. 
thanks: >-
  We thank Mark Buntaine, Michael DeCrescenzo, Jeffrey Friedman, Thomas Leeper, and Tristan Mahr for their helpful comments and input. We also thank Samantha Camilletti and Joshua Dutro for their excellent research assistance. This experiment received approval by the human subjects research committees at both Christopher Newport University (018.042) and Brigham Young University (E18104). We preregistered our hypotheses and research design at the Open Science Framework, and our preregistration protocol is available at <https://osf.io/dx973/>.
author-note: >-
  Suparna Chaudhry, PhD, Assistant Professor of Political Science, Christopher Newport University; Andrew Heiss, PhD, Visiting Assistant Professor, Romney Institute of Public Service and Ethics, Brigham Young University. 
additional-info: Figures should appear in color online only, not in print.
correspondence: >-
  Correspondence concerning this article should be addressed to Suparna Chaudhry, PhD, Department of Political Science, Christopher Newport University, Newport News, VA 23606, 757-594-7821. Contact: suparna.chaudhry@cnu.edu
word-count: "3,998"
---

@@

The 2011 Egyptian uprising presented an opportunity for domestic non-governmental organizations (NGOs) and civil society actors to transform Egypt into a democratic state. However, just as NGOs began to take advantage of new levels of civic freedoms, successive Egyptian governments passed laws restricting these groups from receiving foreign funding or engaging in political advocacy. Egypt is not alone in its non-violent crackdown on NGOs. India, Bolivia, Hungary, China, Russia, and others have imposed funding and programmatic restrictions on NGOs  [@Carothers:2015; @DPR:2015]. These laws have tangible consequences for NGOs worldwide. Official aid flows have decreased substantially in countries that have imposed restrictive anti-NGO laws [@ChaudhryHeiss:2018; @DupuyPrakash:2018], and large foundations and government agencies have both reduced their contributions to NGOs in restrictive countries [@Brechenmacher:2017]. In response to this dramatic shift in the NGO funding landscape, organizations have increasingly turned to individual private donors to offset the loss of governmental and foundation funding [@BanksHulmeEdwards:2015].

Donations from *domestic* donors in repressive countries are unreliable and inconsistent. Citizens in the Global South often cannot support local NGOs, may lack a culture of philanthropic giving, or may channel funds to groups working on non-contentious issues such as health and education [@DPR:2015; @RPC:2016; @Brechenmacher:2017]. Even when donors in repressive countries prefer to donate to NGOs working in contentious areas, they may be deterred by dismal tax incentives, threats to take away business permits, or a general fear of retribution [@HudsonInstitute:2015]. NGOs have increasingly sought out funds from *foreign* individual donors in countries with institutions more amenable to philanthropy. Further, given exchange rates and wealth differentials, it is easier for local NGOs to raise modest amounts of money abroad, rather than investing resources to raise similar sums locally [@RonKairePandya:2018].

Given increased crackdown on NGOs across the world, how do individual donors in the U.S. feel about donating to legally besieged NGOs abroad? Do legal restrictions on NGOs influence donors' decision to donate? We argue that domestic political environments of NGO host countries can influence preferences of foreign private donors and that legal crackdowns on NGOs serve as a heuristic to donors of organizational deservingness. We use a survey experiment to measure the effect of crackdowns on stated donor behavior and find that while crackdowns do not directly influence the likelihood of donation on their own, respondents indicated willingness to donate substantially more to legally restricted NGOs. This effect persists when mediated by other structural heuristics like NGO issue area and funding sources.

# Theory and hypotheses

@@

We propose that the regulatory relationship between international NGOs and their host governments serves as a signal of NGO trustworthiness and deservingness. NGO legislation is not inherently restrictive, and governments frequently use regulations to benignly routinize the behavior of NGOs [@BloodgoodTB]. Restrictive NGO laws, on the other hand, are designed to limit organizational programming by imposing barriers to entry, funding, and advocacy. These stricter anti-NGO laws can signal to donors that NGOs undertake crucial work abroad, precipitating government crackdown. Therefore, we expect that government restrictions abroad will increase respondents' likelihood of donating to the organization and increase the amount donated:

@@

@fig:causal-path represents this relationship more formally. Legal crackdowns (**C**) have a direct effect on donor outcomes (**Y**; donation likelihood and amount) because of their heuristic function. Legal crackdowns do not occur in isolation, however. Countries are more likely to legally restrict NGOs when organizational missions (**I**) are in tension in with government preferences and when organizations receive substantial funding (**F**) from abroad [@DPR:2015]. Each of these organizational characteristics increase the likelihood of crackdown but also simultaneously serve as donor heuristics and influence perceptions of NGO deservingness.

@@

The *source of NGO funding* can also influence donor behavior. Government grants to NGOs can often crowd out private funding [@AndreoniPayne:2003]. Donors may feel less inclined to contribute to NGOs receiving government funding as they might perceive that their tax money already funds the organization. Government funding can also signal a lack of independence from the government. Organizations that receive substantial government funding frequently avoid programming that would question the government even if it is in line with their mission [@Stroup:2012]. In contrast, donors who see that organizations are privately funded may feel that they can also contribute and help.

As funding restrictions are one of the most common forms of anti-NGO regulations [@DPR:2015], organizational funding influences anti-NGO crackdowns (see @fig:causal-path). As such, the combination of legal crackdown and funding may reshape donor perceptions of deservingness. Donors could punish government-funded NGOs that face restrictions, since legal trouble could further fuel private donor mistrust. Donors thus are more likely to donate more to NGOs that rely on private funding:

@@

# Methods

@@

Without considering issue or funding, we find that respondents are not more likely to donate to legally besieged international NGOs (H~1a~). As seen in @fig:likely-diffs(A), those exposed to the crackdown condition tend to have a slightly higher probability of donating (`r percent_1(filter(likely, title == "Crackdown")$theta_2)`) than those in the control condition (`r percent_1(filter(likely, title == "Crackdown")$theta_1)`), but the difference is not significant (`r filter(likely, title == "Crackdown")$delta`; `r filter(likely, title == "Crackdown")$p_neq0`). We find similar results when looking at the isolated effects of issue area and funding. Both differences are positive, but the posterior probability of being larger than zero is low (`r filter(likely, title == "Issue")$delta`; `r filter(likely, title == "Issue")$p_neq0`; `r filter(likely, title == "Funding")$delta`; `r filter(likely, title == "Funding")$p_neq0`). We thus do not initially find evidence to support our hypothesis that crackdowns (or issue or funding) alone increase the likelihood of donating.

@@

@fig:likely-diffs(B–C) shows the difference in donation likelihood for both crackdown conditions across NGO issue areas (H~2a~) and funding sources (H~3a~). Crackdown has almost no effect on the likelihood of donating to human rights NGOs (`r filter(likely, title == "Human rights")$p_neq0`). We hypothesized that the positive crackdown effect would be stronger for humanitarian NGOs, and while crackdown does have a positive effect on under the humanitarian frame, the probability this difference is greater than zero is below our 95% threshold (`r filter(likely, title == "Humanitarian assistance")$p_neq0`). We do find some evidence for H~3a~, however. Crackdown has little effect on the likelihood of donating to NGOs receiving government funding (`r filter(likely, title == "Government")$p_neq0`), but it does increase the likelihood of donating to privately funded NGOs by `r percent_0(filter(likely, title == "Private")$pct_change)` (`r filter(likely, title == "Private")$delta`; `r filter(likely, title == "Private")$p_neq0`).

@@

This study makes two important contributions. While recent work has looked at how state repression of NGOs affects official aid [@ChaudhryHeiss:2018; @DupuyPrakash:2018], there is little research on how repression affects private aid. Further, prior literature on the determinants of donation to NGOs predominantly focuses on individual-level factors. Our study measures important structural determinants that influence giving and shows how these factors interact with organizational attributes to signal NGO deservingness. Additional research should determine the extent to which these preferences translate into behavior—here, we have only asked respondents to make a hypothetical donation—as well as test the generalizability of our findings to other issue areas.

This has several implications for policy and future research. NGOs may benefit from publicizing when they are targets of government crackdown. While this may not be enough to sway the average U.S. citizen, our results show that communicating this information can convince already-likely donors to donate more to besieged groups. Our study also highlights the importance of framing NGO operations. Previous research on human rights micromobilization has shown that the way people process information can affect their increased willingness to help others in need [@Slovic:2007]. Our results show that donors are likely to view legally besieged human rights NGOs with suspicion. However, human rights groups undertake a variety of programs, and communicating the nature of their work and specifying the main target of the organization's efforts can allay the public's fears. Finally, if NGOs signal to donors that receive a majority of their funding from non-governmental sources, they may be able to convince average donors that their contributions might help the NGO resist the crackdown. Moreover, since private funders typically tend to be less interested in the domestic politics of recipient countries compared to governmental funders, relying on private donations is less likely to attract the ire of authoritarian governments fearful of foreign intervention.

>>>>

---
title: "Charity During Crackdown: Analyzing the Impact of State Repression of NGOs on Philanthropy"
short-title: Charity During Crackdown
author:
- name: Suparna Chaudhry
  affiliation: Christopher Newport University
  email: suparna.chaudhry@cnu.edu
  url: http://www.suparnachaudhry.com/
  twitter: SuparnaChaudhry
- name: Andrew Heiss
  affiliation: Brigham Young University
  email: andrew_heiss@byu.edu
  url: https://www.andrewheiss.com/
  twitter: andrewheiss
date: "`r format(Sys.time(), '%B %e, %Y')`"
published: Please do not cite without permission.
git-repo: https://github.com/andrewheiss/ngo-crackdowns-philanthropy
reference-section-title: References
toc: false
title-page: true
abstract: >-
  State crackdown on non-governmental organizations (NGOs), particularly through legal restrictions on funding and advocacy, has become increasingly pervasive across the globe. How do legal restrictions on NGOs impact patterns of private philanthropic giving and individual donor decision-making? With reduced funding for NGOs working on contentious issues, and an absence of philanthropic culture in various developing countries, many NGOs rely on funds from private donors in Western countries. Existing research has predominantly focused on individual-level determinants of philanthropy. However, we argue that domestic political environments of NGO host countries can also influence preferences of foreign private donors, and that crackdown on NGOs serve as a heuristic to donors of organizational deservingness. Using a survey experiment, we explore how this crackdown changes donors' preferences based on the issue area and funding sources of the NGO. We find that while crackdowns do not directly influence the likelihood of donation on their own, donors are willing to donate substantially more to legally besieged NGOs abroad. We conclude by discussing the implications of our results on the sustainability of civil society organizations working in repressive countries. 
thanks: >-
  We thank Mark Buntaine, Michael DeCrescenzo, Jeffrey Friedman, Thomas Leeper, and Tristan Mahr for their helpful comments and input. We also thank Samantha Camilletti and Joshua Dutro for their excellent research assistance. This experiment received approval by the human subjects research committees at both Christopher Newport University (018.042) and Brigham Young University (E18104). We preregistered our hypotheses and research design at the Open Science Framework, and our preregistration protocol is available at <https://osf.io/dx973/>.
author-note: >-
  Suparna Chaudhry, PhD, Assistant Professor of Political Science, Christopher Newport University; Andrew Heiss, PhD, Visiting Assistant Professor, Romney Institute of Public Service and Ethics, Brigham Young University. 
additional-info: Figures should appear in color online only, not in print.
correspondence: >-
  Correspondence concerning this article should be addressed to Suparna Chaudhry, PhD, Department of Political Science, Christopher Newport University, Newport News, VA 23606, 757-594-7821. Contact: suparna.chaudhry@cnu.edu
word-count: "4,081"
---

@@

The 2011 Egyptian uprising presented an opportunity for domestic non-governmental organizations (NGOs) and civil society actors to transform Egypt into a democratic state. However, just as NGOs began to take advantage of new levels of civic freedoms, successive Egyptian governments passed laws restricting these groups from receiving foreign funding or engaging in political advocacy. Egypt is not alone in its non-violent crackdown on NGOs. India, Bolivia, Hungary, China, Russia, and others have imposed funding and programmatic restrictions on NGOs  [@Carothers:2015; @DPR:2015]. These laws have tangible consequences for NGOs worldwide. Official aid flows have decreased substantially in countries that have imposed restrictive anti-NGO laws [@ChaudhryHeiss:2018; @DupuyPrakash:2018], and large foundations and government agencies have both reduced their contributions to NGOs in repressive countries [@Brechenmacher:2017].

In response to this dramatic shift in the NGO funding landscape, organizations have increasingly turned to *foreign* individual donors to offset the loss of governmental and foundation funding [@BanksHulmeEdwards:2015]. This is because donations from *domestic* donors in repressive countries are unreliable and inconsistent. Citizens in the Global South often cannot support local NGOs, may lack a culture of philanthropic giving, or may channel funds to groups working on non-contentious issues such as health and education [@DPR:2015; @RPC:2016; @Brechenmacher:2017]. Even when donors in repressive countries prefer to donate to NGOs working in contentious areas, they may be deterred by dismal tax incentives, threats to take away business permits, or a general fear of retribution [@HudsonInstitute:2015]. Therefore, NGOs have increasingly sought out funds from *foreign* individual donors in countries with institutions more amenable to philanthropy. Further, given exchange rates and wealth differentials, it is easier for local NGOs to raise modest amounts of money abroad, rather than investing resources to raise similar sums locally [@RonKairePandya:2018].

Given increased crackdown on NGOs across the world, how do individual donors in the U.S. feel about donating to legally besieged NGOs abroad? Do legal restrictions on NGOs influence donors' decision to donate? We argue that domestic political environments of NGO host countries can influence preferences of foreign private donors and that legal crackdowns on NGOs serve as a heuristic to donors of organizational deservingness. We use a survey experiment to measure the effect of crackdowns on stated donor behavior and find that while crackdowns do not directly influence the likelihood of donation on their own, respondents indicated willingness to donate substantially more to legally restricted NGOs. This effect persists when mediated by other structural heuristics like NGO issue area and its main funding sources.

# Theorizing the effects of crackdown on philanthropy

@@

We propose that the regulatory relationship between international NGOs and their host governments serves as a signal of NGO trustworthiness and deservingness. NGO legislation is not inherently restrictive, and governments frequently use regulations to benignly routinize the behavior of NGOs [@BloodgoodTB]. Restrictive NGO laws, on the other hand, are designed to limit organizational programming by imposing barriers to entry, funding, and advocacy. These stricter anti-NGO laws can signal to donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and crack down on them. Therefore, we expect that government restrictions abroad will increase respondents' likelihood of donating to the organization and increase the amount donated:

@@

@fig:causal-path represents this relationship more formally. Legal crackdowns (**C**) have a direct effect on donor outcomes (**Y**; donation likelihood and amount) because of their heuristic function. Legal crackdowns do not occur in isolation, however. Countries are more likely to repress NGOs when organizational missions (**I**) are in tension in with government preferences and when organizations receive substantial funding (**F**) from abroad [@DPR:2015]. Each of these organizational characteristics increase the likelihood of crackdown but also simultaneously serve as donor heuristics and influence perceptions of NGO deservingness.

@@

The *source of NGO funding* can also influence donor behavior. Government grants to NGOs can often crowd out private funding [@AndreoniPayne:2003]. Donors may feel less inclined to contribute to NGOs receiving government funding, as they might perceive that their tax money already funds the organization. Government funding can also signal a lack of independence from the government. Organizations that receive substantial government funding frequently avoid programming that would question the government even if it were in line with their mission [@Stroup:2012]. In contrast, donors who see that organizations are privately funded may feel that they can also contribute and help.

As funding restrictions are one of the most common forms of anti-NGO regulations [@DPR:2015], organizational funding can influence anti-NGO crackdowns (see @fig:causal-path). As such, the combination of legal crackdown and funding may reshape donor perceptions of deservingness. Private donors may be more likely to punish government-funded NGOs that face restrictions. This is because legal trouble with the group's main funder could fuel private donor mistrust. Donors thus may be more likely to donate more to NGOs that rely on private funding:

@@

# Research Design

@@

Without considering issue or funding, we find that respondents are not more likely to donate to legally besieged international NGOs (H~1a~). As seen in @fig:likely-diffs(A), those exposed to the crackdown condition tend to have a slightly higher probability of donating (`r percent_1(filter(likely, title == "Crackdown")$theta_2)`) than those in the control condition (`r percent_1(filter(likely, title == "Crackdown")$theta_1)`), but the difference is not significant (`r filter(likely, title == "Crackdown")$delta`; `r filter(likely, title == "Crackdown")$p_neq0`). We find similar results when looking at the isolated effects of issue area and funding. Both differences are positive, but the posterior probability of being larger than zero is low (`r filter(likely, title == "Issue")$delta`; `r filter(likely, title == "Issue")$p_neq0`; `r filter(likely, title == "Funding")$delta`; `r filter(likely, title == "Funding")$p_neq0`). We thus do not initially find evidence that crackdowns (or issue or funding) alone increase the likelihood of donating.

@@

@fig:likely-diffs(B–C) shows the difference in donation likelihood for both crackdown conditions across NGO issue areas (H~2a~) and funding sources (H~3a~). Crackdown has almost no effect on the likelihood of donating to human rights NGOs (`r filter(likely, title == "Human rights")$p_neq0`). We hypothesized that the positive crackdown effect would be stronger for humanitarian NGOs, and while crackdown does have a positive effect under the humanitarian frame, the probability this difference is greater than zero is below our 95% threshold (`r filter(likely, title == "Humanitarian assistance")$p_neq0`). We do find some evidence for H~3a~, however. Crackdown has little effect on the likelihood of donating to NGOs receiving government funding (`r filter(likely, title == "Government")$p_neq0`), but it does increase the likelihood of donating to privately funded NGOs by `r percent_0(filter(likely, title == "Private")$pct_change)` (`r filter(likely, title == "Private")$delta`; `r filter(likely, title == "Private")$p_neq0`).

@@

This study makes two important contributions. While recent work has looked at how state repression of NGOs affects official aid [@ChaudhryHeiss:2018; @DupuyPrakash:2018], there is little research on how repression affects private aid. Further, prior literature on the determinants of donation to NGOs focuses predominantly on individual-level factors. Our study measures important structural determinants that influence philanthropy and shows how these factors interact with organizational attributes to signal NGO deservingness. Additional research should determine the extent to which these preferences translate into behavior—here, we have only asked respondents to make a hypothetical donation—as well as test the generalizability of our findings to other issue areas.

This has several implications for policy and future research. NGOs may benefit from publicizing when they are targets of government crackdown. While this may not be enough to sway the average U.S. citizen, our results show that communicating this information can convince already-likely donors to donate more to besieged groups. Our study also highlights the importance of framing NGO operations. Previous research on human rights micro-mobilization has shown that the way people process information can affect their increased willingness to help others in need [@Slovic:2007]. Our results show that donors are likely to view legally besieged human rights NGOs with suspicion. However, human rights groups undertake a variety of programs, and communicating the nature of their work and specifying the main target of the organization's efforts can allay the public's fears. Finally, if NGOs signal to donors that receive a majority of their funding from non-governmental sources, they may be able to convince average donors that their contributions might help the NGO resist the crackdown. Moreover, since private funders typically tend to be less interested in the domestic politics of recipient countries compared to governmental funders, relying on private donations is less likely to attract the ire of authoritarian governments fearful of foreign intervention.
